DNA Methylation Hierarchical Variance Model\nPaper Figures

### 2021-01-06 Paper_Figures.Rmd
Fixed log2 issue, updated readme

prepdatafolder = "data_final_rg3"
cat(sprintf("Load prepped data dir: %s (needs to be pre-processed)...",prepdatafolder))
gened_data <- read_csv(here(prepdatafolder, "gened_data.csv"), col_types = list("CHR" = col_character()))
prepped_data <- read_csv(here(prepdatafolder, "prepped_data.csv"))
permed_data <- read_csv(here(prepdatafolder, "permed_data.csv"))

@@

\subsection{Prior Probabilities (Fig S1)}
Priors for each parameter (mu, nu, sigmaP/PT/T) were defined using TCGA 450k array data from colorectal tumor patients with paired multiple tumor and/or normal+tumor samples. \newline Empirical values of mean and standard deviations were assesed for all 450k CpG sites. \newline The resulting distributions were fit with: bimodal gaussian (mu), Cauchy (betaT), or gamma distributions (sigmas). \newline
SigmaE (error variance) prior was defined using a non-informative gamma distribution: gamma(2,2).
\newline Additionally: Prior relaxation was performed by algebraically increasing the variance of the distributions (or each Gaussian component in the case of mu) while maintaining the same mode (peak).

```{r Priors, echo=FALSE, fig.fullwidth=TRUE, warning=FALSE, comment=NA}
# load TCGA estimates for mu, betaT, sigmaP, sigmaPT, sigmaT
load(file="/Users/kevinmurgas/Documents/Data+ project/TCGA data/estimates_allpats.Rdata")

# remove all zeros, and take square root of variance (stored variable) to get sigma/stdev
sigmaP_ests <- sqrt(sigmaP_ests[which(!!sigmaP_ests)])
sigmaPT_ests <- sqrt(sigmaPT_ests[which(!!sigmaPT_ests)])
sigmaT_ests <- sqrt(sigmaT_ests[which(!!sigmaT_ests)])
mu_ests <- mu_ests[which(!!mu_ests)]
betaT_ests <- betaT_ests[which(!!betaT_ests)]

@@

# plot TCGA estimates for mu,betaT,sigmaE,sigmaP,sigmaPT,sigmaT; with prior fits overlaid
pM <- post_prior_histogram(mu_ests, function(x) fBG(x,0.3557,-3.0675, 0.75214,1.1290, 1.39197), function(x) fBG(x,0.3557,-3.0675, sqrt(3)*0.75214,1.1290, sqrt(3)*1.39197), 0.1, c(-5,5), "Mu")
pB <- post_prior_histogram(betaT_ests, function(x) dcauchy(x,-0.04051, 0.19330), function(x) dcauchy(x,-0.04051, 0.5799), 0.1, c(-5,5), "Nu")
pP <- post_prior_histogram(sigmaP_ests, function(x) dgamma(x, 5.1989, 16.6192), function(x) dgamma(x, 2.7693, 7.0029), 0.05, c(0,2), "SigmaP")
pPT <- post_prior_histogram(sigmaPT_ests, function(x) dgamma(x, 3.5539, 5.5324), function(x) dgamma(x, 2.1457, 2.4819), 0.05, c(0,2), "SigmaPT")
pT <- post_prior_histogram(sigmaT_ests, function(x) dgamma(x, 6.8383, 25.7560), function(x) dgamma(x, 3.3643, 10.430), 0.05, c(0,2), "SigmaT",legFlag=TRUE)
pE <- ggplot() + stat_function(fun=function(x) dgamma(x, 2, 2),geom="line",color="blue") + ggtitle("SigmaE") + xlim(c(0,2)) + theme_classic() + theme(legend.position = "none") + xlab("SigmaE") + ylab("Freq.")

@@

p_all <- ggarrange(pM,pB,pE,pP,pPT,pT,hjust=0,nrow=2,ncol=3)
print(p_all)
rm(pM,pB,pBlank,pP,pPT,pT,pE)

@@

# plot hierarchical model posterior medians for mu,betaT,sigmaE,sigmaP,sigmaPT,sigmaT; with priors overlaid
pM <- post_prior_histogram(mu_full$p50, genemask, function(x) 1.5*fBG(x,0.3557,-3.0675, 0.75214,1.1290, 1.39197), 0.1, c(-5,5), "Mu")
pB <- post_prior_histogram(betaT_full$p50, genemask, function(x) 1.7*dcauchy(x,-0.04051, 0.5799), 0.1, c(-5,5), "BetaT")
pE <- post_prior_histogram(sigmaE_full$p50, genemask, function(x) 4*dgamma(x,2,2), 0.05, c(0,2), "SigmaE")
pP <- post_prior_histogram(sigmaP_full$p50, genemask, function(x) 2*dgamma(x, 2.7693, 7.0029), 0.05, c(0,2), "SigmaP")
pPT <- post_prior_histogram(sigmaPT_full$p50, genemask, function(x) 1.8*dgamma(x, 2.1457, 2.4819), 0.05, c(0,2), "SigmaPT")
pT <- post_prior_histogram(sigmaT_full$p50, genemask, function(x) 1.2*dgamma(x, 3.3643, 10.430), 0.05, c(0,2), "SigmaT")
# log sigmas
pLP <- post_prior_histogram(log(sigmaP_full$p50), genemask, function(x) 0.52*dgamma(exp(x), 2.7693, 7.0029), 0.1, c(-5,5), "log(SigmaP)")
pLPT <- post_prior_histogram(log(sigmaPT_full$p50), genemask, function(x) 0.7*dgamma(exp(x), 2.1457, 2.4819), 0.1, c(-5,5), "log(SigmaPT)")
pLT <- post_prior_histogram(log(sigmaT_full$p50), genemask, function(x) 0.4*dgamma(exp(x), 3.3643, 10.430), 0.1, c(-5,5), "log(SigmaT)") + theme(legend.position = c(0.8, 0.6), legend.text=element_text(size=6), legend.title=element_text(size=6))

# arrange each parameter as a subplot # ,labels=c("M","B","E","P","PT","T","lP","lPT","lT")
p_all <- ggarrange(pM,pB,pE,pP,pPT,pT,pLP,pLPT,pLT,hjust=0,nrow=3,ncol=3)
print(p_all)
rm(pM,pB,pE,pP,pPT,pT,pLP,pLPT,pLT)
```

\newpage
\subsection{Summary of neff and Rhat for each parameter}
```{r neffRhat, echo=FALSE, warning=FALSE}
# for each parameter (mu, betaT, sigmaP, sigmaPT, sigmaT, sigmaE, lp__)
# reporting: mean n_eff, min n_eff, # n_eff<500, %, # n_eff<100, %
n_eff_mat <-
  data.frame(
  mu = mu_full$n_eff,
  betaT = betaT_full$n_eff,
  sigmaP = sigmaP_full$n_eff,
  sigmaPT = sigmaPT_full$n_eff,
  sigmaT = sigmaT_full$n_eff,
  sigmaE = sigmaE_full$n_eff,
  lp = lp_full$n_eff
  ) %>% gather(param,value,mu:lp) %>% group_by(param) %>% summarize(
  mean = mean(value),
  min = min(value),
  s500 = sum(value < 500),
  p500 = 100 * mean(value < 500),
  s100 = sum(value < 100),
  p100 = 100 * mean(value < 100),
  .groups = "drop"
  ) %>% arrange(match(.$param,c("mu","betaT","sigmaP","sigmaPT","sigmaT","sigmaE","lp")))

@@

# again for Rhat
Rhat_mat <-
  data.frame(
  mu = mu_full$Rhat,
  betaT = betaT_full$Rhat,
  sigmaP = sigmaP_full$Rhat,
  sigmaPT = sigmaPT_full$Rhat,
  sigmaT = sigmaT_full$Rhat,
  sigmaE = sigmaE_full$Rhat,
  lp = lp_full$Rhat
  ) %>% gather(param,value,mu:lp) %>% group_by(param) %>% summarize(
  mean = mean(value),
  max = max(value),
  s11 = sum(value > 1.1),
  p11 = 100 * mean(value > 1.1),
  s101 = sum(value > 1.01),
  p101 = 100 * mean(value > 1.01),
  .groups = "drop"
  ) %>% arrange(match(.$param,c("mu","betaT","sigmaP","sigmaPT","sigmaT","sigmaE","lp")))

@@

# plot the two conservations score prob sums (using /log(2) to convert log_e to log_2)
genemask = !(FullAnnotation$UCSC_RefGene_Name=="")
p_scores <- post_histogram(CpGscore_full$median/log(2),genemask,0.05,"CpG Conservation score") +
  theme(aspect.ratio=1/4,legend.position = "none") +
  scale_x_continuous(limits=c(-4,4),breaks=c(-4,-2,0,2,4)) + 
  scale_y_continuous(breaks=c(0,0.1,0.2,0.3,0.4,0.5))
print(p_scores)
# ggsave("CpGscores_overlay.svg")

# Fig 4B: plot gene scores and adjusted bootstrap distribution
p1 <- permed_data %>% ggplot(aes(x=score_median/log(2))) + geom_histogram(aes(y=..count..), binwidth=0.1, fill="lightgray",color = "grey50") + xlim(c(-1.5,3.5)) + theme_classic() + ylab("Number of genes") + theme(aspect.ratio = 1)
p2 <- permed_data %>% ggplot(aes(x=score_median_v3)) + geom_histogram(aes(y=..count..), binwidth=0.02, fill="lightgray", color="grey50") + xlim(c(0,1)) + theme_classic() + ylab("Number of genes") + theme(aspect.ratio = 1)
p_genes <- ggarrange(p1,p2,nrows=2,ncol=1)
# ggsave("gene_scores_bootstrap.svg")
```

@@

# for a table summary of sigmas and conservation score
prepped_data %>%
  group_by(`Island relation` = Relation_to_UCSC_CpG_Island) %>%
  mutate(score_median = score_median / log(2) ) %>% # convert log_e to log_2
  summarise_at(vars(sigmaP:sigmaT,score_median), mean, na.rm = TRUE) %>%
  dplyr::rename(ir = `Island relation`) %>%
  mutate(ir = case_when(is.na(ir) ~ "Sea",
                        str_detect(ir, "^S") ~ str_replace(ir, "^S_", "South "),
                        str_detect(ir, "^N") ~ str_replace(ir, "^N_", "North "),
                        TRUE ~ ir)) %>%
  dplyr::rename(`Island relation` = ir) %>%
  arrange(c(1,4,2,5,3,6)) %>% # for island, N shore, S shore, N shelf, S shelf, sea
  rename(`CpG Conservation Score`=score_median) %>%
  gt() %>%
  fmt_number(columns = vars(sigmaP, sigmaPT, sigmaT, `CpG Conservation Score`), decimals = 3) %>%
  tab_spanner(label = "Hierarchical Variances", columns = vars(sigmaP, sigmaPT, sigmaT)) %>% cols_align("center")

# stats for cpg island relations
my_data = prepped_data %>% select(score_median, Relation_to_UCSC_CpG_Island) %>%
  mutate(score_median = score_median / log(2) ) %>% # convert log_e to log_2
  group_by(ir = Relation_to_UCSC_CpG_Island) %>%
  mutate(ir = case_when(is.na(ir) ~ "Sea",
                        str_detect(ir, "^S") ~ str_replace(ir, "^S_", ""),
                        str_detect(ir, "^N") ~ str_replace(ir, "^N_", ""),
                        TRUE ~ ir))
# summarize mean of each group and convert to fold change
my_data %>% summarise_at(vars(score_median), mean, na.rm = TRUE) %>% mutate(fc = 2^score_median)
# run Kruskal Wallis ANOVA and post-hoc Tukey
kruskal.test(score_median ~ ir, data = my_data)
#install.packages("MultNonParam")
library(MultNonParam)
tukey.kruskal.test(my_data$score_median, my_data$ir, alpha = 0.001)

# Fig 3B: plot CpG island relation score distributions
prepped_data %>% select(Name, Relation_to_UCSC_CpG_Island, score_median) %>%
  mutate(score_median = score_median / log(2) ) %>% # convert log_e to log_2
  dplyr::rename(ir = Relation_to_UCSC_CpG_Island, score=score_median) %>%
  mutate(ir = case_when(is.na(ir) ~ "Sea",
                        str_detect(ir, "^S") ~ str_replace(ir, "^S_", ""),
                        str_detect(ir, "^N") ~ str_replace(ir, "^N_", ""),
                        TRUE ~ ir)) %>%
  mutate(ir = factor(ir, levels = c("Island", "Shore", "Shelf", "Sea"))) %>%
  #mutate(ir = factor(ir, levels = c("Sea", "Shelf", "Shore", "Island"))) %>%
  ggplot() + geom_violin(aes(x=ir,y=score,fill=ir),draw_quantiles = 0.5) + theme_classic() + scale_y_continuous(limits=c(-5,5),breaks=c(-4,-2,0,2,4)) + theme(aspect.ratio=1/4,legend.position = "none")
# ggsave("CpGscore_islands_violin.svg")
```

@@

temp <- bind_rows({prepped_data %>% summarize_conservation("any") %>% mutate(region="Gene-Associated")},
          {prepped_data %>% summarize_conservation("TSS1500") %>% mutate(region="TSS1500")},
          {prepped_data %>% summarize_conservation("TSS200") %>% mutate(region="TSS200")},
          {prepped_data %>% summarize_conservation("5'UTR|5UTR|5URT") %>% mutate(region="5'UTR")},
          {prepped_data %>% summarize_conservation("1stExon") %>% mutate(region="1stExon")},
          {prepped_data %>% summarize_conservation("ExonBnd") %>% mutate(region="ExonBnd")},
          {prepped_data %>% summarize_conservation("Body") %>% mutate(region="Body")},
          {prepped_data %>% summarize_conservation("3'UTR|3UTR") %>% mutate(region="3'UTR")}) %>%
  select(c(region,cons_in,cons_out,nsites,nout,pval)) %>%
  mutate(cons_in = cons_in/log(2), cons_out = cons_out/log(2)) %>% # change of base e to base 2
  mutate(fold_change = 2^cons_in, fold_diff = 2^(cons_in - cons_out)) # calculating fold difference (2^ (log2 difference))
temp %>% gt() %>% fmt_number(columns = vars(cons_in,cons_out,pval), decimals = 3)

# multiple regression stats (using log2 score)
FRdata = data.frame(CpGscore = log2(prepped_data$sigmaP / prepped_data$sigmaT) ) %>%
  mutate(FR = pull(prepped_data,UCSC_RefGene_Group)) %>%
  mutate(FR=if_else(is.na(FR),"NA",FR)) %>%
  mutate(iTSS200 = str_detect(FR,"TSS200"),
         i5UTR = str_detect(FR,"5'UTR|5UTR|5URT"),
         iBody = str_detect(FR,"Body"),
         i3UTR = str_detect(FR,"3'UTR|3UTR"),
         iTSS1500 = str_detect(FR,"TSS1500"),
         i1stEx = str_detect(FR,"1stExon"),
         iExBnd = str_detect(FR,"ExonBnd"))

@@

# Fig 3C: plot functional region score distributions
temp <- prepped_data %>% select(Name, UCSC_RefGene_Group, score_median) %>%
  mutate(score_median = score_median / log(2) ) %>% # convert log_e to log_2
  separate_rows(UCSC_RefGene_Group, sep=";") %>%
  distinct() %>%
  dplyr::rename(fr = UCSC_RefGene_Group, score=score_median) %>%
  mutate(fr = case_when(fr=="" ~ "NA",
                        str_detect(fr, "5URT") ~ "5'UTR",
                        str_detect(fr, "3UTR") ~ "3'UTR",
                        TRUE ~ fr)) %>%
  distinct()
#keepregions = c("TSS1500","Body", "3'UTR","1stExon","TSS200","5'UTR","ExonBnd")
keepregions = c("TSS200","1stExon", "5'UTR","TSS1500","Body","3'UTR","ExonBnd")
temp %>%
  mutate(fr = factor(fr, levels = keepregions)) %>%
  filter(fr %in% keepregions) %>%
  ggplot() + geom_violin(aes(x=fr,y=score,fill=fr), draw_quantiles = 0.5) + theme_classic() + scale_y_continuous(limits=c(-5,5),breaks=c(-4,-2,0,2,4,6)) + theme(aspect.ratio=1/4,legend.position = "none") + geom_hline(yintercept = 0, color="black", alpha=0.2)
# ggsave("CpGscore_regions_violin.svg")
#rm(temp)

@@

# calculate CpG distance based on Ensembl gene start position and direction
distanced_data <- gened_data %>%
  mutate(score_median = score_median / log(2) ) %>% # convert log_e to log_2
  filter(!is.na(UCSC_RefGene_Name)) %>%
  filter(UCSC_RefGene_Name %in% geneKeys) %>%
  mutate(EDBCHR=edbData$SEQNAME[match(UCSC_RefGene_Name,edbData$GENENAME)],
         EDBSTART=edbData$GENESTART[match(UCSC_RefGene_Name,edbData$GENENAME)],
         EDBDIR=edbData$SEQSTRAND[match(UCSC_RefGene_Name,edbData$GENENAME)],
         EDBTYPE=edbData$GENEBIOTYPE[match(UCSC_RefGene_Name,edbData$GENENAME)]
         ) %>%
  filter(CHR==EDBCHR) %>%
  mutate(cpgdist = if_else(EDBDIR==1, MAPINFO-EDBSTART, EDBSTART-MAPINFO))

@@

plotGeneScore <- function(geneStr,norm) {
  CpGscores <- gened_data %>% filter(str_detect(UCSC_RefGene_Name,paste0("^",geneStr,"$"))) %>% pull(score_median)
  # convert log_e to log_2
  ggplot(prepped_data %>% mutate(score_median = score_median/log(2))) + geom_histogram(aes(x=score_median, y=..count../nrow(prepped_data)), binwidth=0.05,fill="lightgray") + ggtitle(sprintf("%s: CpG scores (%i sites)",geneStr, length(CpGscores))) +
    scale_x_continuous(limits=c(-5,5), breaks=c(-4,-2,0,2,4)) +
    scale_y_continuous(limits=c(0,0.03),breaks=c(0,0.01,0.02,0.03)) +
    xlab("CpG conservation score") + ylab("Freq.") + theme_classic() + theme(legend.position = "none") +
    #geom_vline(xintercept=CpGscores, color="red", alpha=0.5, size=0.2) +
    geom_histogram(data=data.frame(score=CpGscores),aes(x=score,y=norm*..count../length(CpGscores)),binwidth=0.05, fill="blue",alpha=0.5) +
    geom_vline(xintercept=mean(CpGscores), color="blue",linetype="dashed", size=1)
}

>>>>

prepdatafolder = "data_final_rg3"
cat(sprintf("Load prepped data dir: %s (needs to be pre-processed in StanAnalysis.Rmd)...",prepdatafolder))
gened_data <- read_csv(here(prepdatafolder, "gened_data.csv"), col_types = list("CHR" = col_character()))
prepped_data <- read_csv(here(prepdatafolder, "prepped_data.csv"))
permed_data <- read_csv(here(prepdatafolder, "permed_data.csv"))

@@

\subsection{Prior Probabilities (Fig S1)}
Priors for each parameter (mu, nu, sigmaP/PT/T) were defined using TCGA 450k array data from colorectal tumor patients with paired multiple tumor and/or normal+tumor samples. \newline Empirical values of mean and standard deviations were assesed for all 450k CpG sites. \newline The resulting distributions were fit with: bimodal gaussian (mu), Cauchy (nu), or gamma distributions (sigmas). \newline
SigmaE (error variance) prior was defined using a non-informative gamma distribution: gamma(2,2).
\newline Additionally: Prior relaxation was performed by algebraically increasing the variance of the distributions (or each Gaussian component in the case of mu) while maintaining the same mode (peak).

```{r Priors, echo=FALSE, fig.fullwidth=TRUE, warning=FALSE, comment=NA}
# load TCGA estimates for mu, nu, sigmaP, sigmaPT, sigmaT
load(file="/Users/kevinmurgas/Documents/Data+ project/TCGA data/estimates_allpats.Rdata")

# remove all zeros, and take square root of variance (stored variable) to get sigma/stdev
sigmaP_ests <- sqrt(sigmaP_ests[which(!!sigmaP_ests)])
sigmaPT_ests <- sqrt(sigmaPT_ests[which(!!sigmaPT_ests)])
sigmaT_ests <- sqrt(sigmaT_ests[which(!!sigmaT_ests)])
mu_ests <- mu_ests[which(!!mu_ests)]
nu_ests <- nu_ests[which(!!nu_ests)]

@@

# plot TCGA estimates for mu,nu,sigmaE,sigmaP,sigmaPT,sigmaT; with prior fits overlaid
pM <- post_prior_histogram(mu_ests, function(x) fBG(x,0.3557,-3.0675, 0.75214,1.1290, 1.39197), function(x) fBG(x,0.3557,-3.0675, sqrt(3)*0.75214,1.1290, sqrt(3)*1.39197), 0.1, c(-5,5), "Mu")
pN <- post_prior_histogram(nu_ests, function(x) dcauchy(x,-0.04051, 0.19330), function(x) dcauchy(x,-0.04051, 0.5799), 0.1, c(-5,5), "Nu")
pP <- post_prior_histogram(sigmaP_ests, function(x) dgamma(x, 5.1989, 16.6192), function(x) dgamma(x, 2.7693, 7.0029), 0.05, c(0,2), "SigmaP")
pPT <- post_prior_histogram(sigmaPT_ests, function(x) dgamma(x, 3.5539, 5.5324), function(x) dgamma(x, 2.1457, 2.4819), 0.05, c(0,2), "SigmaPT")
pT <- post_prior_histogram(sigmaT_ests, function(x) dgamma(x, 6.8383, 25.7560), function(x) dgamma(x, 3.3643, 10.430), 0.05, c(0,2), "SigmaT",legFlag=TRUE)
pE <- ggplot() + stat_function(fun=function(x) dgamma(x, 2, 2),geom="line",color="blue") + ggtitle("SigmaE") + xlim(c(0,2)) + theme_classic() + theme(legend.position = "none") + xlab("SigmaE") + ylab("Freq.")

@@

p_all <- ggarrange(pM,pN,pE,pP,pPT,pT,hjust=0,nrow=2,ncol=3)
print(p_all)
rm(pM,pN,pBlank,pP,pPT,pT,pE)

@@

# plot hierarchical model posterior medians for mu,nu,sigmaE,sigmaP,sigmaPT,sigmaT; with priors overlaid
pM <- post_prior_histogram(mu_full$p50, genemask, function(x) 1.5*fBG(x,0.3557,-3.0675, 0.75214,1.1290, 1.39197), 0.1, c(-5,5), "Mu")
pN <- post_prior_histogram(nu_full$p50, genemask, function(x) 1.7*dcauchy(x,-0.04051, 0.5799), 0.1, c(-5,5), "Nu")
pE <- post_prior_histogram(sigmaE_full$p50, genemask, function(x) 4*dgamma(x,2,2), 0.05, c(0,2), "SigmaE")
pP <- post_prior_histogram(sigmaP_full$p50, genemask, function(x) 2*dgamma(x, 2.7693, 7.0029), 0.05, c(0,2), "SigmaP")
pPT <- post_prior_histogram(sigmaPT_full$p50, genemask, function(x) 1.8*dgamma(x, 2.1457, 2.4819), 0.05, c(0,2), "SigmaPT")
pT <- post_prior_histogram(sigmaT_full$p50, genemask, function(x) 1.2*dgamma(x, 3.3643, 10.430), 0.05, c(0,2), "SigmaT")
# log sigmas
pLP <- post_prior_histogram(log(sigmaP_full$p50), genemask, function(x) 0.52*dgamma(exp(x), 2.7693, 7.0029), 0.1, c(-5,5), "log(SigmaP)")
pLPT <- post_prior_histogram(log(sigmaPT_full$p50), genemask, function(x) 0.7*dgamma(exp(x), 2.1457, 2.4819), 0.1, c(-5,5), "log(SigmaPT)")
pLT <- post_prior_histogram(log(sigmaT_full$p50), genemask, function(x) 0.4*dgamma(exp(x), 3.3643, 10.430), 0.1, c(-5,5), "log(SigmaT)") + theme(legend.position = c(0.8, 0.6), legend.text=element_text(size=6), legend.title=element_text(size=6))

# arrange each parameter as a subplot # ,labels=c("M","B","E","P","PT","T","lP","lPT","lT")
p_all <- ggarrange(pM,pN,pE,pP,pPT,pT,pLP,pLPT,pLT,hjust=0,nrow=3,ncol=3)
print(p_all)
rm(pM,pN,pE,pP,pPT,pT,pLP,pLPT,pLT)
```

\newpage
\subsection{Summary of neff and Rhat for each parameter}
```{r neffRhat, echo=FALSE, warning=FALSE}
# for each parameter (mu, nu, sigmaP, sigmaPT, sigmaT, sigmaE, lp__)
# reporting: mean n_eff, min n_eff, # n_eff<500, %, # n_eff<100, %
n_eff_mat <-
  data.frame(
  mu = mu_full$n_eff,
  nu = nu_full$n_eff,
  sigmaP = sigmaP_full$n_eff,
  sigmaPT = sigmaPT_full$n_eff,
  sigmaT = sigmaT_full$n_eff,
  sigmaE = sigmaE_full$n_eff,
  lp = lp_full$n_eff
  ) %>% gather(param,value,mu:lp) %>% group_by(param) %>% summarize(
  mean = mean(value),
  min = min(value),
  s500 = sum(value < 500),
  p500 = 100 * mean(value < 500),
  s100 = sum(value < 100),
  p100 = 100 * mean(value < 100),
  .groups = "drop"
  ) %>% arrange(match(.$param,c("mu","nu","sigmaP","sigmaPT","sigmaT","sigmaE","lp")))

@@

# again for Rhat
Rhat_mat <-
  data.frame(
  mu = mu_full$Rhat,
  nu = nu_full$Rhat,
  sigmaP = sigmaP_full$Rhat,
  sigmaPT = sigmaPT_full$Rhat,
  sigmaT = sigmaT_full$Rhat,
  sigmaE = sigmaE_full$Rhat,
  lp = lp_full$Rhat
  ) %>% gather(param,value,mu:lp) %>% group_by(param) %>% summarize(
  mean = mean(value),
  max = max(value),
  s11 = sum(value > 1.1),
  p11 = 100 * mean(value > 1.1),
  s101 = sum(value > 1.01),
  p101 = 100 * mean(value > 1.01),
  .groups = "drop"
  ) %>% arrange(match(.$param,c("mu","nu","sigmaP","sigmaPT","sigmaT","sigmaE","lp")))

@@

# plot the two conservations score prob sums
genemask = !(FullAnnotation$UCSC_RefGene_Name=="")
p_scores <- post_histogram(CpGscore_full$median,genemask,0.05,"CpG Conservation score") +
  theme(aspect.ratio=1/4,legend.position = "none") +
  scale_x_continuous(limits=c(-4,4),breaks=c(-4,-2,0,2,4)) + 
  scale_y_continuous(breaks=c(0,0.1,0.2,0.3,0.4,0.5))
print(p_scores)
# ggsave("CpGscores_overlay.svg")

# Fig 4B: plot gene scores and adjusted bootstrap distribution
p1 <- permed_data %>% ggplot(aes(x=score_median)) + geom_histogram(aes(y=..count..), binwidth=0.1, fill="lightgray",color = "grey50") + xlim(c(-1.5,3.5)) + theme_classic() + ylab("Number of genes") + theme(aspect.ratio = 1)
p2 <- permed_data %>% ggplot(aes(x=score_median_v3)) + geom_histogram(aes(y=..count..), binwidth=0.02, fill="lightgray", color="grey50") + xlim(c(0,1)) + theme_classic() + ylab("Number of genes") + theme(aspect.ratio = 1)
p_genes <- ggarrange(p1,p2,nrows=2,ncol=1)
# ggsave("gene_scores_bootstrap.svg")
```

@@

# for a table summary of sigmas and conservation score
prepped_data %>%
  group_by(`Island relation` = Relation_to_UCSC_CpG_Island) %>%
  summarise_at(vars(sigmaP:sigmaT,score_median), mean, na.rm = TRUE) %>%
  dplyr::rename(ir = `Island relation`) %>%
  mutate(ir = case_when(is.na(ir) ~ "Sea",
                        str_detect(ir, "^S") ~ str_replace(ir, "^S_", "South "),
                        str_detect(ir, "^N") ~ str_replace(ir, "^N_", "North "),
                        TRUE ~ ir)) %>%
  dplyr::rename(`Island relation` = ir) %>%
  arrange(c(1,4,2,5,3,6)) %>% # for island, N shore, S shore, N shelf, S shelf, sea
  rename(`CpG Conservation Score`=score_median) %>%
  gt() %>%
  fmt_number(columns = vars(sigmaP, sigmaPT, sigmaT, `CpG Conservation Score`), decimals = 3) %>%
  tab_spanner(label = "Hierarchical Variances", columns = vars(sigmaP, sigmaPT, sigmaT)) %>% cols_align("center")

# stats for cpg island relations
my_data = prepped_data %>% select(score_median, Relation_to_UCSC_CpG_Island) %>%
  group_by(ir = Relation_to_UCSC_CpG_Island) %>%
  mutate(ir = case_when(is.na(ir) ~ "Sea",
                        str_detect(ir, "^S") ~ str_replace(ir, "^S_", ""),
                        str_detect(ir, "^N") ~ str_replace(ir, "^N_", ""),
                        TRUE ~ ir))
# summarize mean of each group and convert to fold change
my_data %>% summarise_at(vars(score_median), mean, na.rm = TRUE) %>% mutate(fc = 2^score_median)
# run Kruskal Wallis ANOVA and post-hoc Tukey
kruskal.test(score_median ~ ir, data = my_data)
#install.packages("MultNonParam")
library(MultNonParam)
tukey.kruskal.test(my_data$score_median, my_data$ir, alpha = 0.001)

# Fig 3B: plot CpG island relation score distributions
prepped_data %>% select(Name, Relation_to_UCSC_CpG_Island, score_median) %>%
  dplyr::rename(ir = Relation_to_UCSC_CpG_Island, score=score_median) %>%
  mutate(ir = case_when(is.na(ir) ~ "Sea",
                        str_detect(ir, "^S") ~ str_replace(ir, "^S_", ""),
                        str_detect(ir, "^N") ~ str_replace(ir, "^N_", ""),
                        TRUE ~ ir)) %>%
  mutate(ir = factor(ir, levels = c("Island", "Shore", "Shelf", "Sea"))) %>%
  #mutate(ir = factor(ir, levels = c("Sea", "Shelf", "Shore", "Island"))) %>%
  ggplot() + geom_violin(aes(x=ir,y=score,fill=ir),draw_quantiles = 0.5) + theme_classic() + scale_y_continuous(limits=c(-5,5),breaks=c(-4,-2,0,2,4)) + theme(aspect.ratio=1/4,legend.position = "none")
# ggsave("CpGscore_islands_violin.svg")
```

@@

temp <- bind_rows({prepped_data %>% summarize_conservation("any") %>% mutate(region="Gene-Associated")},
          {prepped_data %>% summarize_conservation("TSS1500") %>% mutate(region="TSS1500")},
          {prepped_data %>% summarize_conservation("TSS200") %>% mutate(region="TSS200")},
          {prepped_data %>% summarize_conservation("5'UTR|5UTR|5URT") %>% mutate(region="5'UTR")},
          {prepped_data %>% summarize_conservation("1stExon") %>% mutate(region="1stExon")},
          {prepped_data %>% summarize_conservation("ExonBnd") %>% mutate(region="ExonBnd")},
          {prepped_data %>% summarize_conservation("Body") %>% mutate(region="Body")},
          {prepped_data %>% summarize_conservation("3'UTR|3UTR") %>% mutate(region="3'UTR")}) %>%
  select(c(region,cons_in,cons_out,nsites,nout,pval)) %>%
  mutate(fold_change = 2^cons_in, fold_diff = 2^(cons_in - cons_out)) # calculating fold difference (2^ (log2 difference))
temp %>% gt() %>% fmt_number(columns = vars(cons_in,cons_out,pval), decimals = 3)

# multiple regression stats
FRdata = data.frame(CpGscore = prepped_data$score_median ) %>%
  mutate(FR = pull(prepped_data,UCSC_RefGene_Group)) %>%
  mutate(FR=if_else(is.na(FR),"NA",FR)) %>%
  mutate(iTSS200 = str_detect(FR,"TSS200"),
         i5UTR = str_detect(FR,"5'UTR|5UTR|5URT"),
         iBody = str_detect(FR,"Body"),
         i3UTR = str_detect(FR,"3'UTR|3UTR"),
         iTSS1500 = str_detect(FR,"TSS1500"),
         i1stEx = str_detect(FR,"1stExon"),
         iExBnd = str_detect(FR,"ExonBnd"))

@@

# Fig 3C: plot functional region score distributions
temp <- prepped_data %>% select(Name, UCSC_RefGene_Group, score_median) %>%
  separate_rows(UCSC_RefGene_Group, sep=";") %>%
  distinct() %>%
  dplyr::rename(fr = UCSC_RefGene_Group, score=score_median) %>%
  mutate(fr = case_when(fr=="" ~ "NA",
                        str_detect(fr, "5URT") ~ "5'UTR",
                        str_detect(fr, "3UTR") ~ "3'UTR",
                        TRUE ~ fr)) %>%
  distinct()
#keepregions = c("TSS1500","Body", "3'UTR","1stExon","TSS200","5'UTR","ExonBnd")
keepregions = c("TSS200","1stExon", "5'UTR","TSS1500","Body","3'UTR","ExonBnd")
temp %>%
  mutate(fr = factor(fr, levels = keepregions)) %>%
  filter(fr %in% keepregions) %>%
  ggplot() + geom_violin(aes(x=fr,y=score,fill=fr), draw_quantiles = 0.5) + theme_classic() + scale_y_continuous(limits=c(-5,5),breaks=c(-4,-2,0,2,4,6)) + theme(aspect.ratio=1/4,legend.position = "none") + geom_hline(yintercept = 0, color="black", alpha=0.2)
# ggsave("CpGscore_regions_violin.svg")
#rm(temp)

@@

# calculate CpG distance based on Ensembl gene start position and direction
distanced_data <- gened_data %>%
  filter(!is.na(UCSC_RefGene_Name)) %>%
  filter(UCSC_RefGene_Name %in% geneKeys) %>%
  mutate(EDBCHR=edbData$SEQNAME[match(UCSC_RefGene_Name,edbData$GENENAME)],
         EDBSTART=edbData$GENESTART[match(UCSC_RefGene_Name,edbData$GENENAME)],
         EDBDIR=edbData$SEQSTRAND[match(UCSC_RefGene_Name,edbData$GENENAME)],
         EDBTYPE=edbData$GENEBIOTYPE[match(UCSC_RefGene_Name,edbData$GENENAME)]
         ) %>%
  filter(CHR==EDBCHR) %>%
  mutate(cpgdist = if_else(EDBDIR==1, MAPINFO-EDBSTART, EDBSTART-MAPINFO))

@@

plotGeneScore <- function(geneStr,norm) {
  CpGscores <- gened_data %>% filter(str_detect(UCSC_RefGene_Name,paste0("^",geneStr,"$"))) %>% pull(score_median)
  ggplot(prepped_data) + geom_histogram(aes(x=score_median, y=..count../nrow(prepped_data)), binwidth=0.05,fill="lightgray") + ggtitle(sprintf("%s: CpG scores (%i sites)",geneStr, length(CpGscores))) +
    scale_x_continuous(limits=c(-5,5), breaks=c(-4,-2,0,2,4)) +
    scale_y_continuous(limits=c(0,0.03),breaks=c(0,0.01,0.02,0.03)) +
    xlab("CpG conservation score") + ylab("Freq.") + theme_classic() + theme(legend.position = "none") +
    #geom_vline(xintercept=CpGscores, color="red", alpha=0.5, size=0.2) +
    geom_histogram(data=data.frame(score=CpGscores),aes(x=score,y=norm*..count../length(CpGscores)),binwidth=0.05, fill="blue",alpha=0.5) +
    geom_vline(xintercept=mean(CpGscores), color="blue",linetype="dashed", size=1)
}
